Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-05-27 paper/ifo-singleTone.tex
a few more edits to the draft

The intensity of the interference pattern is measured at a single point on the screen using a commercial USB webcam placed to view the screen (see Fig.~\ref{fig:ifo_schematic_webcam}). 
The webcam's sample rate is $30\,{\rm Hz}$, which limits the spectral content of observable signals to below $15\,{\rm Hz}$, the Nyquist frequency.
A tone under $15\,{\rm Hz}$ is played through the speaker for a minute and the interference pattern recorded with the webcam.

@@

As an initial inspection, we play the timeseries of the intensity using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
The two tones sound similar by ear and don't produce any noticeable beating. 
The result is quantified in Fig.~\ref{fig:webcam_spectrum}. 
A tone of $2.09\,{\rm Hz}$ is played through the speaker and the data recorded via webcam. 
Figure~\ref{fig:webcam_spectrum} shows the Fourier transform of the recording. 
We measure a peak amplitude at $(2.099\pm 0.008)\,{\rm Hz}$ \han{[James: I added in that this is the $1\sigma$ uncertainty, is that correct?] (where the error represents the $1\sigma$ uncertainty).}
The demonstration is able to recover a single frequency tone to within two standard deviations. 
%We played this time series of intensity over time as an audio recording using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
%The two tones played together sound similar and don’t produce any noticeable beating.
%To quantify the precision, we recorded a signal from an injected tone of $2.09\,{\rm Hz}$. 
%From the Fourier transform shown in Figure~\ref{fig:webcam_spectrum}, we measured a peak amplitude at $(2.099\pm 0.008)\,{\rm Hz}$, a result within two standard deviations. 
%The experiment is able to recover a single tone confidently. 
%This showed that the experiment observes a single tone confidently.

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Single injected tone results. 
The Fourier amplitude of the intensity timeseres recorded by webcam. 
The frequency of the injection is $2.09\,{\rm Hz}$. 
The recovered frequency peak is at $(2.099\pm 0.008)\,{\rm Hz}$. 
%The peak (with amplitude $17.5$) has a frequency was measured at $(2.099\pm 0.008) \mathrm{Hz}$. The actual maximum is achieved at $0\,{\rm Hz}$, but corresponds to the signal having a non-zero average value, so we don’t display or consider it.
}

>>>>

The intensity of the interference pattern is measured at a single point on the screen using a commercial USB webcam placed to view the screen (see Fig.~\ref{fig:ifo_schematic_webcam}). 
The webcam's sample rate is $30\,{\rm Hz}$, which limits the spectral content of observable signals to below $15\,{\rm Hz}$, the Nyquist frequency.
A tone under $15\,{\rm Hz}$ is played through the speaker for one minute and the interference pattern recorded with the webcam.

@@

As an initial inspection, we play the timeseries of the intensity using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
The two tones sound similar by ear and don't produce any noticeable beating. 
The result is quantified in Fig.~\ref{fig:webcam_spectrum}. 
A tone of $2.09\,{\rm Hz}$ is played through the speaker and the data recorded via webcam. 
Figure~\ref{fig:webcam_spectrum} shows the Fourier transform of the recording.
\han{[to do - rewrite once James has the new numbers (also Fig 3 caption)] 
We measure a peak amplitude at $(2.099\pm 0.008)\,{\rm Hz}$ (where the error represents the $1\sigma$ uncertainty).
The demonstration is able to recover a single frequency tone to within two standard deviations. }
%We played this time series of intensity over time as an audio recording using the \textbf{io.wavfile.write} function from the Scipy~\cite{scipy} package in Python~\cite{python}. 
%The two tones played together sound similar and don’t produce any noticeable beating.
%To quantify the precision, we recorded a signal from an injected tone of $2.09\,{\rm Hz}$. 
%From the Fourier transform shown in Figure~\ref{fig:webcam_spectrum}, we measured a peak amplitude at $(2.099\pm 0.008)\,{\rm Hz}$, a result within two standard deviations. 
%The experiment is able to recover a single tone confidently. 
%This showed that the experiment observes a single tone confidently.

\begin{figure}
	\includegraphics[width=.49\textwidth]{figures/webcam_expt_4_0209-cropped.pdf}
	\caption{\label{fig:webcam_spectrum}
Single injected tone results. 
The Fourier amplitude of the intensity timeseres recorded by webcam. 
The frequency of the injection is $2.09\,{\rm Hz}$. 
The recovered frequency peak is at $(2.099\pm 0.008)\,{\rm Hz}$. 
\han{[to do: add new numbers from James]}
%The peak (with amplitude $17.5$) has a frequency was measured at $(2.099\pm 0.008) \mathrm{Hz}$. The actual maximum is achieved at $0\,{\rm Hz}$, but corresponds to the signal having a non-zero average value, so we don’t display or consider it.
}

### 2020-05-27 paper/ifo-tableTopGWs.tex
a few more edits to the draft

The Michelson interferometer configuration used in this work is shown schematically in Fig.~\ref{fig:ifo_schematic_webcam}.
It is constructed on a $450 \times 450 \,{\rm mm} $ optical breadboard. 
The laser (top-left) has a wavelength of $532\,{\rm nm}$.
In the configuration shown in Fig.~\ref{fig:ifo_schematic_webcam} the laser beam is first reflected from a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard). 
The beam passes through a a converging lens and then to the beamsplitter, which is $50$/$50$ reflection and transmission. 
The beams are reflected by mirrors at the end of each of the two arms (\han{$\approx XX\,{\rm cm}$ and $\approx XX\,{\rm cm}$ long}) and are recombined at the beamsplitter. 
The resulting interference pattern is projected onto a screen and recorded by a webcam.

>>>>

The Michelson interferometer configuration used in this work is shown schematically in Fig.~\ref{fig:ifo_schematic_webcam}.
It is constructed on a $450 \times 450 \,{\rm mm} $ optical breadboard. 
The laser (top-left) has a wavelength of $532\,{\rm nm}$.
In the configuration shown in Fig.~\ref{fig:ifo_schematic_webcam} the laser beam is first reflected from a mirror which turns the beam $90^{\circ}$ (to save space on the optical breadboard). 
The beam passes through a a converging lens and then to the beamsplitter, which is $50$/$50$ reflection and transmission. 
The beams are reflected by mirrors at the end of each of the two arms ($\approx 7.5\,{\rm cm}$ and $\approx 10.0\,{\rm cm}$ long) and are recombined at the beamsplitter. 
The resulting interference pattern is projected onto a screen and recorded by a webcam.

### 2020-05-27 paper/ifo-wanderingTone.tex
a few more edits to the draft

The analysis techniques used here are inspired by those used to search for continuous waves from spinning neutron in LIGO and Virgo data.
Continuous wave searches are performed on long datasets (months to years in duration). 
The frequency of gravitational-wave emission wanders significantly over the observation period.

One method to search for a wandering signal is to split the data into several shorter segments which are analysed seperately initially. 
The resulting segments are combined to form a grid in time and frequency (spectogram). 
The value at each spectrogram point determins the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave analysis, a detection statistic is used to calculate the likelihood of a signal (the statistic used depends on the type of target\han{refs}). 
In this work we take a Fourier transfprm of each time segment.
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin. 
The task is then to find a wandering frequency signal through the spectogram. 
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method find the best path through the grid. 
It is used in contiuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

In section~\ref{sec:viterbi} we review the Viterbi algorithm and its application to this work and in section~\ref{sec:wanderingResults} we present our results for recovery of a wandering audio frequency.

@@

\subsection{The Viterbi algorithm}
\label{sec:viterbi}
%Here we overview the Viterbi algorithm and describe its usage in this work.

Abstractly speaking, the algorithm is given a weighted graph (e.g.\ the spectrogram grid), a sensible sequence of subgraphs (e.g.\ the columns of the spectrum at each time), and restrictions on connectivity (e.g.\ the allowed amount frequency wander). 
At each iteration, the algorithm will find the best path to each node in the next sub-graph. 
This is repeated to the final sub-graph. 
%n, at each iteration, it will find the best path to each node in the next subgraph, all the way from the first subgraph. 
At the end, it selects the overall best path from the first to the last subgraph (here, from the start to end time), this overall best path is called the Viterbi path. 
See Appendix~\ref{app:viterbi} for the full details of the algorithm in this particular implementation.

% moved from above
The method requires a limit for the allowed frequency change over time, usually informed by a physical model. 
Here, the frequency wander limit is set to $0.11\,{\rm Hz}$ which corresponds to one frequency bin per every time bin. 
In some cases, we see this proves poorly adjusted to the signal.

@@

Our results are shown in Figure~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pick-dot and white-cross markers show the injected signal and recovered Viterbi path respectively. 
To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
The Viterbi path stays within one frequency bin (approximately $0.3\,{\rm Hz}$) of the injected tone for $94\%$ or the run. \han{to check}
Note that the signal initially moves faster than the frequency wander limit, causing a discrepancy between the injected and recovered signals. 
To fix this, the limit should be doubled, although this would then cause the recovered path to stray more often. 
There also appears to be an anomaly at $150\,{\rm s}$, likely from a disturbance (such as someone walking) around the interferometer.

>>>>

The analysis techniques used here are inspired by those used to search for continuous waves from spinning neutron in LIGO and Virgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}.
Continuous wave searches are performed on long datasets (months to years in duration). 
The frequency of gravitational-wave emission wanders significantly over the observation period.

One method to search for a wandering signal is to split the data into several shorter segments which are analysed separately initially. 
The resulting segments are combined to form a grid in time and frequency (spectogram). 
The value at each spectrogram point determines the likelihood that a signal is present at a particular time and frequency. 
In gravitational-wave analysis, a detection statistic is used to calculate the likelihood of a signal (the statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}). 
In this work we take a Fourier transform of each time segment.
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin. 
The task is then to find a wandering frequency signal through the spectogram.

The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence. 
It is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

In Section~\ref{sec:viterbi} we review the Viterbi algorithm and its application to this work and in Section~\ref{sec:wanderingResults} we present our results for recovery of a wandering audio frequency.

@@

\subsection{The Viterbi algorithm}
\label{sec:viterbi}
%Here we overview the Viterbi algorithm and describe its usage in this work. 
%Thinking about this - should we add some maths?
%Our starting point is the spectrogram grid in time $t$ and frequency $f$. 
%Each element in the grid contains the Fourier amplitude for a particular $f$ and $t$. 
%Our objective is to find the most likely through the grid which maximises the summed Fourier amplitudes along the path. 
%In our implementation we sum the log of the normalised 
%We label the time-frequency grid points as $t_i$ and $f_j$ respectively, where $i=1,2,...N_t$ and $j=1,2,...,N_f$ (and $N_t$ and $N_f$ are the total number of time and frequency bins respectively). 
%Starting at $t_{i=2}$, the algorithm iterates through $t_i$, at each stage finding the best path to every $f_j$ from $t_{i=0}$.

Abstractly speaking, the algorithm is given a weighted graph (e.g.\ the spectrogram grid), a sensible sequence of subgraphs (e.g.\ the columns of the spectrum at each time), and restrictions on connectivity (e.g.\ the allowed amount of frequency wander). 
At each iteration, the algorithm will find the best path to each node in the next sub-graph. 
This is repeated to the final sub-graph. 
%n, at each iteration, it will find the best path to each node in the next subgraph, all the way from the first subgraph. 
The Viterbi path is then the overall best path from the first sub-graph to the last. 
The overall best path from the first sub-graph to the last is then selected, which is called the Viterbi path. 
At the end, it selects the overall best path from the first to the last subgraph (here, from the start to end time), this overall best path is called the Viterbi path. 
See Appendix~\ref{app:viterbi} for the full details of the algorithm in this particular implementation.

% moved from above
The method requires a limit for the allowed frequency change over time, usually informed by a physical model. 
In this work the frequency $f$ path can move in one of three ways at each time bin: a) up one frequency bin; b) stay in the same frequency bin; or c) down one frequency bin. 
We set the frequency bin width to be $\Delta f = 0.3\,{\rm Hz}$ which corresponds to a maximum frequency wander of $0.11\,{\rm Hz}{\rm s}^{-1}$.
%In this work, the frequency wander limit is set to $0.11\,{\rm Hz} {\rm s}^{-1}$ which corresponds to one frequency bin per every time bin. 
In some cases, we see this proves poorly adjusted to the signal.

@@

Our results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pick-dot and white-cross markers show the injected signal and recovered Viterbi path respectively. 
The Viterbi path provides a good match to the injected frequency. 
%To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
%The Viterbi path stays within one frequency bin (approximately $0.3\,{\rm Hz}$) of the injected tone for $94\%$ of the run. 
Initially the injected signal moves faster then the frequency wander limit, leading to the discrepancy for $t\lesssim 10\,{\rm s}$. 
%Note that the signal initially moves faster than the frequency wander limit, causing a discrepancy between the injected and recovered signals. 
%To fix this, the limit should be doubled, although this would then cause the recovered path to stray more often. 
There is also an anomaly at $150\,{\rm s}$, which is likely due to a disturbance around the interferometer (such as someone walking nearby).
%There also appears to be an anomaly at $150\,{\rm s}$, likely from a disturbance (such as someone walking) around the interferometer.
